---
title: "Converting `sp` Spatial Data Frames to GeoJSON"
author: M. Edward (Ed) Borasky
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why would we do this?
There are two reasons we might want to do this:

1. A user doesn't have R and thus can't do anything with an `sp` data frame, or
2. We want to store the data frame in a database, and the database doesn't have a way to read an R data file.

## Why GeoJSON?

1. It's a standard open source format.
2. It can be easily read by `QGIS` and many JavaScript mapping tools.
3. It can be uploaded to a PostGIS database.

## What libraries do we need?
```{r}
library(sp)
library(readr)
library(rgdal)

```

## Read the R data frame
```{r}
bus_stop_catchment_zone <- readr::read_rds("/Raw/busstop_catchment_zone_shp.RDS")

```

## Write the GeoJSON
```{r}
rgdal::writeOGR(bus_stop_catchment_zone, 
                dsn = "/Work/bus_stop_catchment_zone.geojson", 
                layer = "bus_stop_catchment_zone", 
                driver = "GeoJSON",
                overwrite_layer = TRUE
)

```

